Aligning sequence reads, clone sequences and assembly contigs with BWA-MEM

### 2013-05-26 bwamem.tex
removed a vspace{}
better formatting in arXiv; change to abstract

\begin{abstract}
\section{Summary:} BWA-MEM is a new alignment algorithm for aligning sequence
reads or assembly contigs against a large reference genome such as human.
It automatically chooses between local and end-to-end alignments, supports
paired-end reads and reports multiple local hits. The algorithm is robust to
sequencing errors and applicable to a wide range of sequence lengths from 70bp
to a few megabases. For mapping 100bp sequences, BWA-MEM shows better
performance than several state-of-art read aligners to date.

@@

In this background, a few long-read alignment algorithms, including
BWA-SW~\citep{Li:2010fk}, Bowtie2~\citep{Langmead:2012fk},
Cushaw2~\citep{Liu:2012uq} and GEM~\citep{Marco-Sola:2012kx}, have been
developed. However, they all have some weakness. BWA-SW is slower than bowtie2
for 100bp reads at a comparable accuracy and less accurate then Cushaw2 at a
comparable speed. Bowtie2 and Cushaw2 are slower for 600bp reads (see Results).
While GEM is both fast and accurate for up to approximately 1000bp reads, it
mandates end-to-end alignment and does not perform affine-gap alignment, which
limits its uses for long-read alignment. Even for typical sequence reads ranged
from 100bp to 1000bp in length, no mappers work well across the full spectrum.
At the same time, the increasing read length not only calls for new alignment
algorithms, but also opens the opportunity to de novo assembly which typically
results in contigs ranged from several hundred base pairs to a few megabases.
Very few algorithms are able to align sequences of such variable lengths
at high accuracy and to be robust to translocations in the assembly. All these
concerns motivated us to explore a new alignment algorithm.

\vspace{-1em}

>>>>

\begin{abstract}
\section{Summary:} BWA-MEM is a new alignment algorithm for aligning sequence
reads or assembly contigs against a large reference genome such as human.
It automatically chooses between local and end-to-end alignments, supports
paired-end reads and performs chimeric alignment. The algorithm is robust to
sequencing errors and applicable to a wide range of sequence lengths from 70bp
to a few megabases. For mapping 100bp sequences, BWA-MEM shows better
performance than several state-of-art read aligners to date.

@@

In this background, a few long-read alignment algorithms, notably including
BWA-SW~\citep{Li:2010fk}, Bowtie2 \citep{Langmead:2012fk},
Cushaw2~\citep{Liu:2012uq} and GEM~\citep{Marco-Sola:2012kx}, have been
developed. However, they all have some weakness. BWA-SW is slower than bowtie2
for 100bp reads at a comparable accuracy and less accurate then Cushaw2 at a
comparable speed. Bowtie2 and Cushaw2 are slower for 600bp reads (see Results).
While GEM is both fast and accurate for up to approximately 1000bp reads, it
mandates end-to-end alignment and does not perform affine-gap alignment, which
limits its uses for long-read alignment. Even for typical sequence reads ranged
from 100bp to 1000bp in length, no mappers work well across the full spectrum.
At the same time, the increasing read length not only calls for new alignment
algorithms, but also opens the opportunity to de novo assembly which typically
results in contigs ranged from several hundred base pairs to a few megabases.
Very few algorithms are able to align sequences of such variable lengths
at high accuracy and to be robust to translocations in the assembly. All these
concerns motivated us to explore a new alignment algorithm.
